---
title: "Practical Machine Learning Course Project"
author: "Bruce Brown"
date: "4/26/2017"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background (from the course website)

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The csv files have been downloaded to the working directory.

## Purpose and Process.

The goal of this project is to predict the manner in which participants did the exercise. This is the "classe" variable. This report will describe how a Random Forest model was used for prediction, The prediction model will then be used to predict 20 different test cases as required by the assignment. 

## Required Libraries.

```{r}
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
```


## Downloading and Inspecting the Data.

Read the csv files.

```{r}
trainSet <- read.csv("~/Desktop/MachineLearning/PML Project/pml-training.csv")
testSet <- read.csv("~/Desktop/MachineLearning/PML Project/pml-testing.csv")
```

```{r}
dim(trainSet)
dim(testSet)
```

The training set has 19622 observations of 160 variables.
The test set has 20 observations of 160 variables.

Inspect the variables.
```{r}
names(trainSet)
```

## Cleaning the Data.

Remove variables that contain missing values.
```{r}
trainSet <- trainSet[, colSums(is.na(trainSet)) == 0] 
testSet <- testSet[, colSums(is.na(testSet)) == 0] 

```
Remove variables that have near zero variance.

```{r}
nzv <- nearZeroVar(trainSet)
trainSet <- trainSet[, -nzv]

```
```{r}
nzv <- nearZeroVar(testSet)
testSet <- testSet[, -nzv]

```
Remove variables that may not contribute to prediction. It appears from their labels that the first 7 columns are variables that may not assist in predictive measurement.

```{r}
trainSet <- trainSet[, -(1:7)]
testSet  <- testSet[, -(1:7)]

dim(trainSet)
dim(testSet)
```
Cleaning the data reults in 52 variables to be used for prediction.

##Training the Model.

Partition trianSet data into 2 parts, 60% and 40%. The reulting 60% dataTrain is used to train the model. The 40% dataTest is used to validate the model. Seed is set for reproducibility.

```{r}
set.seed(11111)

inTrain <- createDataPartition(trainSet$classe, p=0.6, list = FALSE)

dataTrain <- trainSet[inTrain,]
dataTest <- trainSet[-inTrain,]

dim(dataTrain)
dim(dataTest)
```
Create the model on dataTrain using the Random Forest Function.
```{r}
modFit <- randomForest(classe~., data = dataTrain)
print(modFit)
```
Apply the fitted model on dataTest for cross validation.

```{r}
predictA <- predict(modFit, dataTest, type = "class")
confusionMatrix(dataTest$classe, predictA)
```
Apply the fitted model on dataTrain.

```{r}
predictB <- predict(modFit, dataTrain, type = "class")
confusionMatrix(dataTrain$classe, predictB)
```
Apply the fitted model to testSet to predict the 20 test cases.
```{r}
predict_TEST <- predict(modFit, testSet, type = "class")
print(predict_TEST)
```
##Result.
The fitted model generated by the Random Forest function resulted in an estimated error rate of 0.59%. When applied to the testing set (dataTest) the accuracy rate was calculated to be 0.9931. The fitted model successfully predicted all 20 test cases. 
